---
title: "Environmental & species data"
subtitle: "Climate, soil, landscape, taxonomy, phylogeny & functional"
author: "Sylvain Schmitt"
institute: CIRAD
date: today
date-format: medium
format: 
  revealjs:
    theme: dark
    output-location: fragment
    slide-number: true
    logo: figures/logo.png
    transition: fade
    preview-links: true
    chalkboard: true
bibliography: references.bib
include-in-header:
  - text: |
      <style>
      .reveal .slide-logo {
        max-height: unset;
        height: 75px;
      }
      </style>
---

```{r set}
#| message: false
#| include: false
library(tidyverse)
library(sf)
library(leaflet)
library(units)
library(ggdark)
library(ggfortify)
```

# Environment

**environment** is a sub-project of the [**BioForest**](https://github.com/Bioforest-project) project aimed at gathering environmental variables of climate ([TerraClimate](https://www.climatologylab.org/terraclimate.html)), soil ([SoilGrids](https://www.isric.org/explore/soilgrids)) and landscape ([TMF](https://forobs.jrc.ec.europa.eu/TMF)) for data preparation within the project.

## Sites

```{r map}
#| message: false
#| warning: false
#| fig-cap: "Sites and plots locations."
read_tsv("../environment/data/derived_data/sites.tsv") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  leaflet() %>%
  addTiles() %>%
  addProviderTiles("Esri.WorldImagery", group = "ESRI") %>%
  addMarkers(
    label = ~ paste(site, plot),
    labelOptions = labelOptions(noHide = FALSE)
  )
```

## Surface

```{r sitesarea}
#| message: false
#| warning: false
#| fig-cap: "Number of sites and total plots surface."
read_tsv("../environment/data/derived_data/sites.tsv") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  group_by(site) %>%
  summarise(n = n()) %>%
  st_convex_hull() %>%
  mutate(area = as.numeric(set_units(st_area(.), "km^2"))) %>%
  st_drop_geometry() %>%
  ggplot(aes(n, area)) +
  geom_point() +
  xlab("Number of plots") +
  ylab(expression(paste("Area [", km^2, "]"))) +
  ggrepel::geom_text_repel(aes(label = site), color = "white") +
  dark_theme_gray() +
  theme(plot.background = element_rect(fill = "grey10", color = NA)) +
  scale_y_sqrt()
```

# Climate

[TerraClimate](https://www.climatologylab.org/terraclimate.html) is a dataset of **monthly** & **\~4-km (1/24th degree**)climate and climatic **water balance** for global terrestrial surfaces from **1958-2019** [@abatzoglou2018]. We have information per depth in each plot as shown below, but I focuses on the top horizon (up to 5cm).

## Monthly variables

-   ET: Actual evapotranspiration, mm
-   CWD: Climate water deficit, mm
-   Tmax: Maximum temperature, °C
-   Tmin: Minimum temperature, °C
-   PDSI: Palmer Drought Severity Index
-   PET: Potential evapotranpiration, mm
-   Pr: precipitation, mm
-   Soil: Soil humidity, mm
-   VPD: Vapour Pressure Deficit, kPa

## Ombrothermal

```{r umbrothermal}
#| message: false
#| warning: false
#| fig-cap: "Umbrothermal diagrams for the mean climate of each site."
read_tsv("../environment/data/derived_data/climate_month.tsv") %>%
  filter(site != "Uppangala") %>%
  group_by(site, month) %>%
  summarise_all(mean) %>%
  ggplot(aes(x = month)) +
  geom_col(aes(y = pr / 10, fill = pr < 100), col = NA) +
  geom_line(aes(y = tmmn, col = "maximum"), linewidth = 1.1) +
  geom_line(aes(y = tmmx, col = "minimum"), linewidth = 1.1) +
  xlab("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
  ) +
  scale_y_continuous(
    name = "Temperature (°C)",
    sec.axis = sec_axis(trans = ~ . * 10, name = "Precipitation (mm)")
  ) +
  scale_color_discrete(guide = "none") +
  facet_wrap(~site) +
  scale_fill_manual(guide = "none", values = c("darkgrey", "white")) +
  dark_theme_gray() +
  theme(plot.background = element_rect(fill = "grey10", color = NA))
```

## Ombrothermal Uppangala

```{r umbrothermal_uppangala}
#| message: false
#| warning: false
#| fig-cap: "Umbrothermal diagrams for the mean climate of Uppangala."
read_tsv("../environment/data/derived_data/climate_month.tsv") %>%
  filter(site == "Uppangala") %>%
  group_by(site, month) %>%
  summarise_all(mean) %>%
  ggplot(aes(x = month)) +
  geom_col(aes(y = pr / 10, fill = pr < 100), col = NA) +
  geom_line(aes(y = tmmn, col = "maximum"), linewidth = 1.1) +
  geom_line(aes(y = tmmx, col = "minimum"), linewidth = 1.1) +
  xlab("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
  ) +
  scale_y_continuous(
    name = "Temperature (°C)",
    sec.axis = sec_axis(trans = ~ . * 10, name = "Precipitation (mm)")
  ) +
  scale_color_discrete(guide = "none") +
  facet_wrap(~site) +
  scale_fill_manual(guide = "none", values = c("darkgrey", "white")) +
  dark_theme_gray() +
  theme(plot.background = element_rect(fill = "grey10", color = NA))
```

## Yearly metrics {.smaller}

-   CWD: Climate water deficit, mm, mean across years of the sum of monthly CWD
-   ET: Evapotranspiration, mm, mean across years of the sum of monthly ET
-   Pr: Precipitation, mm, mean across years of the sum of monthly Pr
-   Soil: Soil humidity, mm, mean across years of the minimum of monthly soil humidity
-   VPD: Vapour Pressure Deficit, kPa, mean across years of the maximum of monthly VPD
-   Tmax: Maximum temperature, °C, mean across years of the maximum of monthly Tmax
-   DSL: Dry season length, days, mean across years of the number of month with ET\>Pr multiplied by 30
-   DSI: Dry season intensity, mm, mean across years of the sum of ET-Pr for month with ET\>Pr

## Metrics

```{r metricssoil}
#| message: false
#| warning: false
#| fig-cap: "Yearly mean climate variables per site."
read_tsv("../environment/data/derived_data/climate_year.tsv") %>%
  gather(variable, value, -site, -plot) %>%
  mutate(var_long = recode(variable,
    "et" = "Evapotranspiration [ mm ]",
    "cwd" = "Climate water deficit [ mm ]",
    "pr" = "Precipitation [ mm ]",
    "tmax" = "Max temperature [ °C ]",
    "vpd" = "Max VPD [ kPA ]",
    "soil" = "Min soil humidity [ mm ]",
    "dsi" = "Dry season intensity [ mm ]",
    "dsl" = "Dry season length [ day ]"
  )) %>%
  ggplot(aes(site, value, fill = site, group = paste(site, plot))) +
  geom_col(position = "dodge") +
  facet_wrap(~var_long, scales = "free_y") +
  dark_theme_gray() +
  theme(
    legend.position = "right", axis.title = element_blank(),
    axis.text.x = element_blank(),
    legend.background = element_rect(fill = "grey10", color = NA),
    plot.background = element_rect(fill = "grey10", color = NA)
  ) +
  scale_fill_discrete("")
```

## Co-variations

```{r pcasoil}
#| message: false
#| warning: false
#| fig-cap: "Yearly mean climate variables PCA per site and plot."
data <- read_tsv("../environment/data/derived_data/climate_year.tsv")
pca <- prcomp(data %>% select(-site, -plot), scale. = TRUE)
autoplot(pca,
  loadings = TRUE, loadings.label = TRUE,
  loadings.label.repel = TRUE,
  data = data, colour = "site"
) +
  scale_color_discrete("") +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

## Correlations

```{r corrsoil}
#| message: false
#| warning: false
#| fig-cap: "Climate variables pairwise correlations."
par(bg = "grey10")
read_tsv("../environment/data/derived_data/climate_year.tsv") %>%
  select(-site, -plot) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(
    type = "upper", diag = FALSE,
    bg = "grey10", tl.col = "white"
  )
```

# Soil

[SoilGrids](https://www.isric.org/explore/soilgrids) are global **soil property** maps at **six standard depth** intervals (according to the GlobalSoilMap IUSS working group and its specifications) at a spatial resolution of **250 meters** [@poggio2021]. I focused on the **two top horizons (0-15cm)** revealing most inter-site variations (V. Freycon pers. com.).

## Variables {.smaller}

-   Clay: Proportion of clay particles (\< 0.002 mm) in the fine earth fraction, %
-   Sand: Proportion of sand particles (\> 0.05 mm) in the fine earth fraction, %
-   Silt: Proportion of silt particles (≥ 0.002 mm and ≤ 0.05 mm) in the fine earth fraction, %
-   BD: Bulk density of the fine earth fraction, kg dm³
-   CEC: Cation Exchange Capacity of the soil, cmolC kg^-1^
-   CF: Volumetric fraction of coarse fragments (\> 2 mm), cm^3^ 100cm^-3^
-   N: Total nitrogen, g kg^-1^
-   pH: Soil pH
-   SOC: Soil organic carbon content in the fine earth fraction, g kg^-1^
-   OCD: Organic carbon density, kg dm³
-   OCS: Organic carbon stocks, kg m^2^

## Paracou

```{r par_soil}
#| message: false
#| warning: false
#| fig-cap: "Soil data variation with depth at Paracou across plots."
read_tsv("../environment/data/derived_data/soil.tsv") %>%
  filter(site == "Paracou") %>%
  select(-X, -Y, -site, -ocs) %>%
  gather(variable, value, -plot, -depth) %>%
  mutate(var_long = recode(variable,
    "bdod" = '"BD ["~kg~dm^{-3}~"]"',
    "cec" = '"CEC ["~cmol~kg^{-1}~"]"',
    "cfvo" = '"CF ["~cm^3~"100"~cm^{-3}~"]"',
    "clay" = '"Clay ["~"%"~"]"',
    "nitrogen" = '"N ["~g~kg^{-1}~"]"',
    "ocd" = '"OCD ["~kg~dm^{3}~"]"',
    "phh2o" = "pH",
    "sand" = '"Sand ["~"%"~"]"',
    "silt" = '"Silt ["~"%"~"]"',
    "soc" = '"SOC ["~g~kg^{-1}~"]"'
  )) %>%
  ggplot(aes(depth, value, group = as.character(plot))) +
  geom_line() +
  facet_wrap(~var_long, scales = "free_y", labeller = label_parsed) +
  ylab("") +
  xlab("Depth [ m ]") +
  scale_x_log10() +
  ggtitle("Paracou") +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

## All

```{r allplot5}
#| message: false
#| warning: false
#| fig-cap: "Mean soil data per site for top horizon (up to 5cm)."
read_tsv("../environment/data/derived_data/soil.tsv") %>%
  select(-X, -Y, -plot) %>%
  group_by(site) %>%
  filter(depth <= 15) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  gather(variable, value, -site, -depth) %>%
  na.omit() %>%
  mutate(var_long = recode(variable,
    "bdod" = '"BD ["~kg~dm^{-3}~"]"',
    "cec" = '"CEC ["~cmol~kg^{-1}~"]"',
    "cfvo" = '"CF ["~cm^3~"100"~cm^{-3}~"]"',
    "clay" = '"Clay ["~"%"~"]"',
    "nitrogen" = '"N ["~g~kg^{-1}~"]"',
    "ocd" = '"OCD ["~kg~dm^{3}~"]"',
    "ocs" = '"OCS ["~kg~m^{-2}~"]"',
    "phh2o" = "pH",
    "sand" = '"Sand ["~"%"~"]"',
    "silt" = '"Silt ["~"%"~"]"',
    "soc" = '"SOC ["~g~kg^{-1}~"]"'
  )) %>%
  ggplot(aes(site, value, fill = site)) +
  geom_col(aes(group = as.factor(depth)), position = "dodge") +
  facet_wrap(~var_long, scales = "free_y", labeller = label_parsed) +
  scale_fill_discrete("") +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  ) +
  theme(
    legend.position = "bottom", axis.title = element_blank(),
    axis.text.x = element_blank()
  )
```

## Textures

```{r betterplot}
#| message: false
#| warning: false
#| fig-cap: "Soil texture for all sites, plots and depths."
par(bg = "grey10")
read_tsv("../environment/data/derived_data/soil.tsv") %>%
  ggplot(aes(x = sand, y = clay, z = silt, col = site)) +
  ggtern::coord_tern(L = "x", T = "y", R = "z") +
  geom_point() +
  labs(
    yarrow = "Clay [ % ]",
    zarrow = "Silt [ % ]",
    xarrow = "Sand [ % ]"
  ) +
  ggtern::theme_showarrows() +
  ggtern::theme_hidetitles() +
  ggtern::theme_clockwise() +
  scale_color_discrete("") +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

## Co-variations

```{r pca}
#| message: false
#| warning: false
#| fig-cap: "Soil variables PCA per site and plot."
data <- read_tsv("../environment/data/derived_data/soil.tsv") %>%
  select(-X, -Y, -plot) %>%
  group_by(site) %>%
  filter(depth <= 15) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  select(-ocs) %>%
  na.omit()
pca <- prcomp(data %>% select(-site, -depth), scale. = TRUE)
autoplot(pca,
  loadings = TRUE, loadings.label = TRUE,
  loadings.label.repel = TRUE,
  data = data, colour = "site"
) +
  scale_color_discrete("") +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

## Correlations

```{r corr}
#| message: false
#| warning: false
#| fig-cap: "Soil variables pairwise correlations."
par(bg = "grey10")
read_tsv("../environment/data/derived_data/soil.tsv") %>%
  select(-X, -Y, -plot) %>%
  group_by(site) %>%
  filter(depth <= 15) %>%
  summarise_all(mean, na.omit = TRUE) %>%
  select(-ocs) %>%
  na.omit() %>%
  select(-site, -depth) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(
    type = "upper", diag = FALSE,
    bg = "grey10", tl.col = "white"
  )
```

# Landscape

The European Commission’s Joint Research Centre developed a dataset on **forest cover change** in tropical moist forests [TMF](https://forobs.jrc.ec.europa.eu/TMF) using 42 years of Landsat time series at 0.09 ha resolution (**30m**) depicting disturbances (deforestation and degradation), and post-deforestation recovery (or forest regrowth) over the period 1990-2023 [@europeancommission.jointresearchcentre.2020].

## Trajectories

```{r allsites}
#| message: false
#| warning: false
#| fig-cap: "Forest cover through time."
read_tsv("../environment/data/derived_data/landscape.tsv") %>%
  na.omit() %>%
  ggplot(aes(year, 100 - intact, col = site, group = paste(site, plot))) +
  geom_line() +
  xlab("") +
  ylab("Deforested area in a radius of 1-km [ % ]") +
  facet_wrap(~site) +
  scale_color_discrete(guide = "none") +
  scale_y_sqrt() +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

## Deforested

```{r deforested}
#| message: false
#| warning: false
#| fig-cap: "Deforested area in a radius of 1-km, %."
read_tsv("../environment/data/derived_data/landscape.tsv") %>%
  group_by(site, plot) %>%
  summarise(deforested = mean(100 - intact)) %>%
  ggplot(aes(site, deforested, fill = site, group = paste(site, plot))) +
  geom_col(position = "dodge") +
  xlab("") +
  ylab("Deforested area in a radius of 1-km [ % ]") +
  scale_fill_discrete(" ") +
  coord_flip() +
  scale_y_sqrt() +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

# All environment

## Co-variations

```{r pcaall}
#| message: false
#| warning: false
#| fig-cap: "All variables PCA per site and plot."
data <- read_tsv("../environment/outputs/environment.tsv") %>%
  na.omit()
pca <- prcomp(data %>% select(-site, -plot, -longitude, -latitude),
  scale. = TRUE
)
autoplot(pca,
  loadings = TRUE, loadings.label = TRUE,
  loadings.label.repel = TRUE,
  data = data, colour = "site"
) +
  scale_color_discrete("") +
  dark_theme_gray() +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    legend.background = element_rect(fill = "grey10", color = NA)
  )
```

## Correlations

```{r corrall}
#| message: false
#| warning: false
#| fig-cap: "All variables pairwise correlations."
par(bg = "grey10")
read_tsv("../environment/outputs/environment.tsv") %>%
  select(-site, -plot, -longitude, -latitude) %>%
  cor(use = "pairwise.complete.obs") %>%
  corrplot::corrplot(
    type = "upper", diag = FALSE,
    bg = "grey10", tl.col = "white"
  )
```

# Species

# References

::: {#refs}
:::
